### 2025-04-09 correlation.qmd
20250409 - Anscombe's quartet

The formula for the correlation coefficient\index{correlation} is in @eq-pearsonCorrelation.
As noted in @sec-statisticalTestsPartitionedVariance, the correlation coefficient can be thought of as the ratio of shared variance (i.e., covariance) to total variance, as in @eq-correlationPartitionedVariance.

@@

>>>>

The formula for the ([Pearson](#sec-correlationExamplesPearson)) correlation coefficient\index{correlation} is in @eq-pearsonCorrelation.
As noted in @sec-statisticalTestsPartitionedVariance, the correlation coefficient can be thought of as the ratio of shared variance (i.e., covariance) to total variance, as in @eq-correlationPartitionedVariance.

@@

Keep in mind that the ([Pearson](#sec-correlationExamplesPearson)) correlation examines the strength of the *linear* association between two variables.\index{correlation}
If the association between two variables is nonlinear, the ([Pearson](#sec-correlationExamplesPearson)) correlation provides the strength of the linear trend and may not provide a meaningful index of the strength of the association between the variables.\index{correlation}
For instance, Anscombe's quartet includes four sets of data that have nearly identical basic descriptive statistics (see Tables [-@tbl-anscombe] and [-@tbl-anscombeStats]), including the same bivariate correlation, yet have very different distributions and whose association takes very different forms (see Figure \@ref(fig:anscombeQuartet)).\index{correlation}

| x1 | y1    | x2 | y2   | x3 | y3    | x4 | y4   |
|----|-------|----|------|----|-------|----|------|
| 10 | 8.04  | 10 | 9.14 | 10 | 7.46  | 8  | 6.58 |
| 8  | 6.95  | 8  | 8.14 | 8  | 6.77  | 8  | 5.76 |
| 13 | 7.58  | 13 | 8.74 | 13 | 12.74 | 8  | 7.71 |
| 9  | 8.81  | 9  | 8.77 | 9  | 7.11  | 8  | 8.84 |
| 11 | 8.33  | 11 | 9.26 | 11 | 7.81  | 8  | 8.47 |
| 14 | 9.96  | 14 | 8.1  | 14 | 8.84  | 8  | 7.04 |
| 6  | 7.24  | 6  | 6.13 | 6  | 6.08  | 8  | 5.25 |
| 4  | 4.26  | 4  | 3.1  | 4  | 5.39  | 19 | 12.5 |
| 12 | 10.84 | 12 | 9.13 | 12 | 8.15  | 8  | 5.56 |
| 7  | 4.82  | 7  | 7.26 | 7  | 6.42  | 8  | 7.91 |
| 5  | 5.68  | 5  | 4.74 | 5  | 5.73  | 8  | 6.89 |

: Anscombe's Quartet {#tbl-anscombe}

| Property                     | Value        |
|------------------------------|--------------|
| Sample size                  | 11           |
| Mean of X                    | 9.0          |
| Mean of Y                    | ~7.5         |
| Variance of X                | 11.0         |
| Variance of Y                | ~4.1         |
| Equation of regression line  | Y = 3 + 0.5X |
| Standard error of slope      | 0.118        |
| One-sample t-statistic       | 4.24         |
| Sum of squares of X          | 110.0        |
| Regression sum of squares    | 27.50        |
| Residual sum of squares of Y | 13.75        |
| Correlation coefficient      | .816         |
| Coefficient of determination | .67          |

: Descriptive Statistics of Anscombe's Quartet {#tbl-anscombeStats}

```{r}
#| label: fig-anscombeQuartet
#| fig-cap: "Anscombe's Quartet."
#| fig-alt: "Anscombe's Quartet."
#| code-fold: true

anscombe

par(mfrow = c(2,2))

plot(
  anscombe$x1,
  anscombe$y1,
  xlab = "x",
  ylab = "y",
  xlim = c(0,20),
  ylim = c(0,15),
  main = 
    substitute(
      paste(italic(r), " = ", x, sep = ""),
      list(x = round(cor.test(x = anscombe$x1, y = anscombe$y1)$estimate, 2))))

abline(lm(
  y1 ~ x1,
  data = anscombe),
  col = "black",
  lty = 2)

plot(
  anscombe$x2,
  anscombe$y2,
  xlab = "x",
  ylab = "y",
  xlim = c(0,20),
  ylim = c(0,15),
  main = substitute(
    paste(italic(r), " = ", x, sep = ""),
    list(x = round(cor.test(x = anscombe$x2, y = anscombe$y2)$estimate, 2))))

abline(lm(
  y2 ~ x2,
  data = anscombe),
  col = "black",
  lty = 2)

plot(
  anscombe$x3,
  anscombe$y3,
  xlab = "x",
  ylab = "y",
  xlim = c(0,20),
  ylim = c(0,15),
  main = substitute(
    paste(italic(r), " = ", x, sep = ""),
    list(x = round(cor.test(x = anscombe$x3, y = anscombe$y3)$estimate, 2))))

abline(lm(
  y3 ~ x3,
  data = anscombe),
  col = "black",
  lty = 2)

plot(
  anscombe$x4,
  anscombe$y4,
  xlab = "x",
  ylab = "y",
  xlim = c(0,20),
  ylim = c(0,15),
  main = substitute(
    paste(italic(r), " = ", x, sep = ""),
    list(x = round(cor.test(x = anscombe$x4, y = anscombe$y4)$estimate, 2))))

abline(lm(
  y4 ~ x4,
  data = anscombe),
  col = "black",
  lty = 2)
```
